---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ramiga <a href="https://pepijn-devries.github.io/ramiga/"><img src="man/figures/logo.svg" align="right" height="139" alt="ramiga website" /></a>

<!-- badges: start -->
[![R-CMD-check](https://github.com/pepijn-devries/ramiga/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pepijn-devries/ramiga/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The status of this package is currently highly experimental. It provides R
bindings to the [vAmiga](https://github.com/dirkwhoffmann/vamiga) emulator.

It will:

  * Provide programmatic access to the emulator
  * Emulate Amiga's with the original chip set.
    * So, neither enhanced chip set nor the Advanced Graphics Architecture
      are supported
  * Capture output
  * Provide input (mouse movements, keyboard events, etc.) programmatically

It will **not**:

  * Provide full interactive visuals and peripherals
    * There is simply no way to exchange this information between
      the R environment and the emulator with sufficient speed and efficiency.

## Installation

You can install the development version of ramiga like so:

``` r
remotes::install_github("pepijn-devries/ramiga")
```

## Example

```{r example}
library(ramiga)
emu <- RamigaEmulator$new()

emu$power_on()

```

## Not on CRAN

This package will not be appearing on CRAN any time soon. If this package
were to be submitted to CRAN there are several hurdles to be taken:

  * All code should comply with ISO standards.
    * With some effort this would be achievable
  * Package size should not exceed 5 MB, but should at best be below 10 MB
    * This will be difficult, if we want to keep all features
  * We should avoid using non-standard compilation flags
    (like -Wa,-mbig-obj)
    * This could be fixed by splitting up large objects into multiple
      smaller ones.
  * This package relies heavily on C++20 features. This will
    not fly on all CRAN build and check machines (which is a requirement).
    * This will take way too much effort to convert to CRAN acceptable
      alternatives.

I pay as much attention to quality as any of my other R packages, so
you should be able to enjoy using this package. But for the reasons stated
above, I will not submit to CRAN. You can install it from R-Universe as
specified above.

*Why not use libretro?*
I did explore this avenue. It doesn't need C++20, so in that respect it
would be easier to get on CRAN. However, the library is much bulkier than
vAmiga. It also doesn't have a sleek and easy to implement API like vAmiga.
So I'll leave this challenge for someone else to confront.