---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ramiga <a href="https://pepijn-devries.github.io/ramiga/"><img src="man/figures/logo.svg" align="right" height="139" alt="ramiga website" /></a>

<!-- badges: start -->
![Status: Experimental](https://img.shields.io/badge/status-experimental-orange)
[![R-CMD-check](https://github.com/pepijn-devries/ramiga/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pepijn-devries/ramiga/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The status of this package is currently highly experimental. It provides R
bindings to the [vAmiga](https://github.com/dirkwhoffmann/vamiga) emulator.

It will:

  * Provide programmatic access to the emulator
  * Emulate Amiga's with the original chip set.
    * So, neither enhanced chip set nor the Advanced Graphics Architecture
      are supported
  * Capture output
  * Provide input (mouse movements, keyboard events, etc.) programmatically

It will **not**:

  * Provide full interactive visuals and peripherals
    * There is simply no way to exchange this information between
      the R environment and the emulator with sufficient speed and efficiency.

## Installation

You can install the development version of ramiga like so:

```{r github, eval=FALSE}
remotes::install_github("pepijn-devries/ramiga")
```

If available for your OS, you can install from R-Universe:

```{r r-univers, eval=FALSE}
install.packages("ramiga", repos = c('https://pepijn-devries.r-universe.dev', 'https://cloud.r-project.org'))
```

## Example

In order to get started, you need to create a new RamigaEmulator object.
As the emulator only emulates the Commodore Amiga machine, you
need to provide a ROM image. For this you could use official licensed
Kickstart ROM images, or you use an AROS ROM image, which is released
in the public domain. In the example below we start the machine with
the AROS ROM, for which we also need the extension.

```{r initiate}
library(ramiga)
emu <- RamigaEmulator$new()

emu$memory$load_rom(
  "https://github.com/vAmigaWeb/vAmigaWeb/raw/refs/heads/main/roms/aros-rom-20250219.bin")

emu$memory$load_rom_extension(
  "https://github.com/vAmigaWeb/vAmigaWeb/raw/refs/heads/main/roms/aros-ext-20250219.bin")
```

Next we load a disk image, which represents a virtual floppy disk.
We use the 'State of the Art' demo by Spaceballs as an ultimate
90s test case. We insert the virtual disk in the first floppy
drive of the emulator, then power on the virtual machine.

```{r insert-disk}
disk <- RamigaImage$new(
  "https://ftp.amigascne.org/pub/amiga/Groups/S/Spaceballs/Spaceballs-StateOfTheArt.dms")

emu$floppy_drives[[1]]$insert_disk( disk )

emu$power_on()
```

The emulator starts in a paused state, as we don't want to block the
main R thread. You actively have to tell the emulator to progress.
Let's skip 1,269 frames while the machine boots. They are not that
interesting.

```{r skip, eval=FALSE}
emu$fast_forward(1269L)
```

Now we can progress frame by frame, and capture the video image and
the audio output buffer for each frame. In the example below, we
store the audio in a `matrix`, and the frames as `".png"` files in the
temporary directory.

```{r capture-frames, eval=FALSE}
audio <- matrix(double(), 2, 0)

#drain whatever is in the buffer first:
dummy <- emu$output$capture_audio_buffer()

## Update frame, refill audio buffer with one frame worth of audio:
emu$next_frame()

## vector of frame indices:
idxs <- 1L:1550L
frame_files <- file.path(tempdir(), sprintf("fr%04i.png", idxs))
for (i in idxs) {
  audio <- cbind(audio, emu$output$capture_audio_buffer())
  emu$next_frame()
  emu$output$capture_video_frame( frame_files[[i]] )
}
```

Now save the `matrix` as a `".wav"` file, such that it can be
used for rendering a video. We can also use the audio to calculate
the emulated time span.

```{r save-audio, eval=FALSE}
## Save the samples as a ".wav" such that we can use it for rendering
audio_file  <- tempfile(fileext = ".wav")
save_wav(audio, audio_file)

## The duration is dictated by the number of samples
## divided by the sample rate
sample_rate <- 44100L
duration    <- ncol(audio)/sample_rate
frame_rate  <- length(idxs)/duration
```

If we have the [av](https://docs.ropensci.org/av/) package installed,
we can render the captured frames and audio to a video file.

```{r render-video, eval=FALSE}
if (requireNamespace("av")) {
  video_file  <- tempfile(fileext = ".mp4")
  
  ## We use the 'vfilter' argument to correct the aspect ratio
  ## of the captured video frames.
  av::av_encode_video(
    frame_files, video_file, frame_rate,
    vfilter = "scale=iw:2*ih:flags=neighbor,setdar=4/3",
    audio = audio_file)
}
```

We have now successfully rendered the boot block intro of the demo:

<video width="100%" controls>
  <source src="https://github.com/pepijn-devries/ramiga/raw/refs/heads/main/data-raw/state-of-the-art.mp4" type="video/mp4">
  Your browser does not support the video tag.
</video>

## Not on CRAN

This package will not be appearing on CRAN any time soon. If this package
were to be submitted to CRAN there are several hurdles to be taken:

  * All code should comply with ISO standards.
    * With some effort this would be achievable
  * Package size should not exceed 5 MB, but should at best be below 10 MB
    * This will be difficult, if we want to keep all features
  * We should avoid using non-standard compilation flags
    (like -Wa,-mbig-obj)
    * This could be fixed by splitting up large objects into multiple
      smaller ones.
  * This package relies heavily on C++20 features. C++20 is standardised
    since R 4.6.0, we need to wait until this becomes old-release.
    as CRAN expects compatibility with old- and new-release.

I pay as much attention to quality as any of my other R packages, so
you should be able to enjoy using this package. But for the reasons stated
above, I will not submit to CRAN. You can install it from R-Universe as
specified above.

*Why not use libretro?*
I did explore this avenue. It doesn't need C++20, so in that respect it
would be easier to get on CRAN. However, the library is much bulkier than
vAmiga. It also doesn't have a sleek and easy to implement API like vAmiga.
So I'll leave this challenge for someone else to pick up.